Exploring the conformational dynamics of a large molecule in solution using an external electric field: A nonequilibrium molecular dynamics simulation

### 2013-11-06 writedown/npt-neq/alanine.tex
change according to Giovanni's suggestions. big changes are going to come

The possible effect of the electric field (EF) on the conformation of
proteins is a subject of increasing attention since it is linked to the problem of exposing human tissues to
electromagnetic radiation. Despite its major relevance, a
satisfactory understanding of how the EF influences the behaviour of proteins
has not been reached so far although a considerable number of
experimental~\cite{bohr2000microwave, bohr2000microwave-1,
  dePomerai2000cell, inskip2001cellular, mancinelli2004non} and
theoretical~\cite{budi2005electric, budi2007effect,
  budi2008comparative, toschi2008effects, astrakas2011electric,
  astrakas2012structural, damm2012can, starzyk2013proteins,
  english2009nonequilibrium, solomentsev2012effects} studies have been devoted to this topic;
where the references reported above represent only a minor sample of examples.

In this perspective, molecular dynamics (MD) has been proved to be useful in
understanding the behavior of a protein 
in both static and oscillatory EF~\cite{budi2005electric, budi2007effect, budi2008comparative,
  toschi2008effects, astrakas2011electric, astrakas2012structural,
  damm2012can, starzyk2013proteins, english2009nonequilibrium,
  solomentsev2012effects}. In these studies, the initial
configurations are usually taken from the Protein Data Bank (PDB),
and then temperature controlled simulations are performed for a maximum extension of the order of 10-60 of nanoseconds.
During this simulation time 
relevant conformational changes characterizing the solvated molecule can be observed. 
Along the trajectories, various observable, such as the
secondary structure, root mean square displacement (RMSD), the dipole
moment, shape parameters are calculated as a function of time or taking its time average. This approach releases useful information on how the
secondary structure of the molecule is changed by external  EF.  However,
by plotting some quantities against time, one may observe strong
fluctuations; the entity of such fluctuations are such that a quantitative 
analysis becomes difficult and does not allow for  a clear understanding of the rationale behind the conformational changes.
Although the time average reduces the statistical uncertainty,
however in this specific case, it is indeed not well defined;
in fact we are in the presence of a nonequilibrium process
(e.g. relaxing process to the new conformations under a static
EF, or the dynamics from one conformation to the other under an oscillatory EF).

In this perspective, if one wants to use MD, the proper way to do it would be that of treating a situation of nonequilibrium with algorithms that can explicitly describe it. This is the approach that will be followed in this work.

The method we employ is the dynamical nonequilibrium molecular dynamics simulation (D-NEMD).
The method was firstly developed by one of us (GC) and coworkers~\cite{ciccotti1975direct, ciccotti1979thought}.
It has been recently applied, with success, to the study of
hydrodynamics~\cite{orlandini2011hydrodynamics,
  orlandini2011hydrodynamics-01}. The essential feature of D-NEMD is the following: It provides a way of evaluating
nonequilibrium observables by using a properly determined ensemble of nonequilibrium MD
simulations.  In this context, the method can be used for analyzing the
conformational changes of a solvated molecule, subject to a time-dependent EF.
In the present paper, we study a small peptide, alanine dipeptide, as a model for showing how the method can be used. 
The choice of a small peptide, rather than a larger one, is done firstly because it
is easier to draw clear conclusions on the nonequilibrium properties
of a single peptide by excluding a possible interplay between
different peptide segments along a chain.
Secondly, the computational cost required is rather massive and thus the testing of the various methodology developments proposed here would not be easy to accomplish for large molecules. Of course, given a positive outcome for such a model system, the aim is that of approaching systems of growing complexity.
The paper is organized as follows:
We firstly describe the idea of D-NEMD, then
introduce the physical
observables for analyzing the nonequilibrium properties. The numerical results of alanine
dipeptide under static and sine-shaped oscillatory EF are
finally presented. By calculating the nonequilibrium properties from D-NEMD
simulations, we are able to address the question of how the peptide
changes the conformation in an EF (conformational space), when do these changes happens (time scale),
and how likely they are (probability). 
To our knowledge, this study is one of the few (if not the first)  where nonequilibrium w.r.t. conformation is explicitly considered for a large molecule in solution subject to time-dependent EF.

\section{Methodology}
\subsection{Nonequilibrium molecular dynamics simulation}
In this section we briefly review the approach for performing dynamical nonequilibrium
molecular dynamics (D-NEMD) \cite{ciccotti1975direct, ciccotti1979thought,
  orlandini2011hydrodynamics, orlandini2011hydrodynamics-01}.
In the following we denote the macroscopic observable by $O(t)$. If at $t$
the configurational probability distribution is $\rho(\vect x, t)$, where
$\vect x$ is the phase space variable, then the observable can be
expressed by
\begin{align}\label{eqn:tmp1}
  O(t) = \int d\vect x\, \hat O(\vect x)\rho(\vect x, t)  = \langle \hat O(\vect x), \rho(\vect x, t)\rangle,
\end{align}
we refer to $\hat O (\vect x)$ as the microscopic observable, 
which is measured at each point in the phase space $\vect x$.
Here we always assume that the initial probability distribution
$\rho(\vect x, 0)$ is known and is identical to the equilibrium distribution of the system
without EF.
The bracket on the right hand side of Eq.~\eqref{eqn:tmp1} denotes, as usual, the inner product in the
phase space.  We assume that he dynamics of the system is governed by the
Hamiltonian equation, i.e. $\dot {\vect x} = J \cdot \nabla_{\vect x}
\mathcal H(\vect x)$, where $\mathcal H$ is the Hamiltonian, and $J$ is
the symplectic matrix. The Liouville equation for the probability
distribution writes:
\begin{align}\label{eqn:tmp2}
  \frac{\partial \rho(\vect x, t)}{\partial t} = - iL(t) \rho(\vect x, t),
\end{align}
where $iL(t) = \{\cdot, \mathcal H\}$ is the Liouville operator.
Eq.~\eqref{eqn:tmp2}
can be formally solved by $\rho(\vect x, t) = e^{-iL(t)} \rho(\vect x, 0)$.
On the other hand
\begin{align}
  \frac{d \hat O(\vect x(t))}{dt} = \nabla_{\vect x}\hat O\cdot \dot{\vect x}
  = \nabla_{\vect x}\hat O\cdot J\cdot \nabla_{\vect x}\mathcal H
  = iL(t) \hat O (\vect x(t))
\end{align}
This equation can be formally solved by $\hat O(\vect x(t)) = e^{iL(t)} \hat O(\vect x(0))$, therefore,
\begin{align}\nonumber
  O(t) & = \langle \hat O(\vect x), \rho(\vect x, t)\rangle
  = \langle \hat O(\vect x), e^{-iL(t)} \rho(\vect x, 0)\rangle
  = \langle e^{iL(t)}\hat O(\vect x), \rho(\vect x, 0)\rangle\\\label{eqn:tmp4}
  &= \langle \hat O(\vect x(t)), \rho(\vect x, 0)\rangle
\end{align}
Since we assume that the system starts from the distribution of equilibrium (without EF)
 Eq.~\eqref{eqn:tmp4} expresses the fact that the 
observable $O(t)$ calculated in a situation of nonequilibrium is equal to the ensemble average of
the microscopic observable measured along trajectories, the
initial configurations of which are subject to the \emph{equilibrium}
distribution. In practical terms we proceed by first running an equilibrium MD simulation
to generate a sample of configurations. Next we employ these configurations as
initial configurations and, for each, the Hamiltonian dynamics is integrated until
time $t$. These trajectories are also called \emph{branching
  trajectories} in this paper (see Fig.\ref{branching} for a pictorial representation). 
Finally, the value of the macroscopic observable at time $t$ is
calculated by averaging the microscopic observable measured at each time $t$ along each branching trajectory.
Stochastic dynamics (Langevin) can be hadeled analogoursly.

@@

\subsection{Control of temperature in Nonequilibrium MD}\label{sec:tmp2b}

\begin{figure}
  \centering
  \includegraphics[width=0.3\textwidth]{thermostat.eps}
  \caption{Schematic plot of the nonequilibrium temperature control scheme.
    The white area is the region characterized by Hamiltonian dynamics without the direct action of an external thermostat.
    The gray area is the region where the external thermostat is used in order to keep the target temperature.
    $R_{ex}$ denotes the size of the Hamiltonian dynamics region, which
    is a sphere at the center of the box.
    The whole system is subject to the periodic boundary conditions. }
  \label{fig:tmp2}
\end{figure}

% The aforementioned D-NEMD simulation algorithm
% does not describe how the boundary of the system should be set up.
% The very naive idea of doing the simulation
% is to simulate a infinitely large isolated system, or any large enough
% isolated subsystem of it.
% In practice, 
% Therefore, we are interested in a 
For this study, we are interested in 
nonthermal~\cite{delaHoz2005microwaves} effects of the EF
on the configuration of the solvated alanine dipeptide.
For this reason a proper control of the temperature in a nonequilibrium system
becomes an important issue; in fact it was argued that  an
accurate control of the internal reaction temperature is essential for being able to
perform reproducible experiments~\cite{damm2012can}.
A straightforward (and realistic) implementation of the temperature control consists in 
embedding the alanine into an infinitely large solvent environment, and 
the EF is only applied to the neighborhood of the alanine. In this case
any extra heat generated by the EF can be effectively
dispersed in the large solvent environment.
In this way one can identify the effect of the EF on the conformational space of the solvated molecule,
without the artifact that the surrounding solvent cannot release the additional energy of the EF.
However, in practice, in most of the cases the computational cost of having large solvent environment is rather massive if not prohibitive.
Therefore, it is of interest to substitute the infinitely large
solvent environment with a smaller affordable system and at the same time accurately reproduce the nonequilibrium
physical properties of interest.

As it is well known, in MD, for equilibrium situations, the solution is to couple
the system to a an external thermostat, under which the desired ensemble
(canonical in this case) can be
sampled by simulating a finite size system with periodic boundary
conditions.
However, coupling the system to a thermostat for nonequilibrium situations is a delicate issue.
In fact, since a thermostat is explicitly designed to bring the system into equilibrium, any perturbation produced by the EF would be mostly adsorbed by the thermostat. This means that the physics of the original Hamiltonian dynamics with the EF would be lost and the response to the perturbation becomes unphysical. 
%will disturb the original Hamiltonian dynamics. This is
%in general problematic, 
%because the dynamics  under thermostating is artificial designed,
%so that the equilibrium distribution can be correctly produce, but not
%the nonequilibrium properties of system.
% dynamics of the system is no longer Hamiltonian, and is
% artificial designed to reach the equilibrium probability distribution.
% However, for a nonequilibrium system, the \emph{dynamics} properties
% will be distrubed by the thermostat, because the
% dynamics of the system is no longer Hamiltonian, and is
% artificial designed to reach the equilibrium probability distribution.
% For a nonequilibrium simulation, simply couple the system
% to a thermostat may not be a good idea, because the \emph{dynamics}
% generated by the thermostat is artificial, and deviation from the
% Hamiltonian dynamics under investigation will introduce artificial
% a effects in the nonequilibrium observations.
In this section we propose a procedure to solve the problem: we
observe the Hamiltonian dynamics (without thermostat) of the alanine 
itself and the water molecules in the closer solvation shells is
crucial, in order to determine the response of the observable of interest
under the perturbation of the EF (in this work, it is the conformational change of alanine).
Instead the detailed dynamics of
the water molecules far away is not of primary importance, that is the region beyond the first solvation shells plays the role of a thermodynamic bath only. 
Therefore, instead of simulating an infinitely
large system, we only simulate a finite size system with periodic
boundary condition with the following characteristics. The simulating region is divided in two
subregions: a spherically defined region around the alanine dipeptide, centered at the
alpha-carbon of radius $R_{ex}$. Here the dynamics is
Hamiltonian and molecules are not subject to a thermostat; we refer to such subregion as \emph{the Hamiltonian dynamics
region}. Instead beyond $R_{ex}$, the dynamics of water is coupled
to a Langevin thermostat, and this region is called \emph{the thermostated
region} (see Fig.~\ref{fig:tmp2}). For technical convenience we
fix the position of the alpha-carbon in space, so that the alanine is always
located at the center of the simulation region.
Therefore, a realistic nonequilibrium  dynamics is preserved in the Hamiltonian dynamics region (where the properties of interest are observed)
and possible direct artificial effects of the thermostating process in the region outside are negligible because of the finite correlation
range of liquid water.
At the same time, the thermostated region works as a infinitely large
environment that effectively absorbs the extra heat produced by the EF.
The validity of the above
statement will be later on carefully checked by a series of numerical tests where it is shown that
the response of observables in nonequilibrium (i.e. under the effect of the EF) does not depend on the size of the system
and on the size of the Hamiltonian dynamics subregion, provided that they are
reasonably large.

@@

\section{Example I: Alanine dipeptide
  under a uniform  constant EF}

\subsection{System settings and simulation protocol}
At time $t=0$~ps, the system
has been fully equilibriated without any EF. From $t=0$ to
the warm-up time $t=t_{\textrm{warm}}$, the eclectic field is switched on linearly, while
after $t=t_{\textrm{warm}}$, the field is kept constant in time at
$E_{\infty}$.  In this work we consider $t_{\textrm{warm}} = 10$~ps,
and $E_{\infty} = 1$~V/nm.
The direction of the field is arbitrarily chosen along the
$x$ direction, in fact the dipole of the equilibrium configurations of
alanine can be along any direction.
In this paper we denote the
EF as a function of time $\vect E(t)$.
Therefore, the constant EF is
$\vect E(t) = (E_\infty\cdot t/t_{warm},0,0)$ as $0\leq t < t_{warm}$, and 
$\vect E(t) = (E_\infty,0,0)$
as $t \geq t_{warm}$.
The system is set up in a $2.7\times 2.7\times
2.7\, \textsf{nm}^3$ periodic simulation region, with one alanine dipeptide
described by the CHARMM27 force field, and solved in 644 TIP3P
water molecules.
%The EF exserts force $\vect F = q \vect E(t)$ on
%the partial charge $q$ in the system.
The size of the  Hamiltonian dynamics region is $R_{ex} = 1.0$~nm.
All simulations are performed by a home-modified GROMACS 4.5~\cite{pronk2013gromacs}.
First, an equilibrium NVT simulation of
100~\textsf{ns} was performed with a Langevin thermostat (time-scale
$\tau_T = 0.5~\textsf{ps}$).  Along the trajectory, configurations were taken every
50~\textsf{ps} and we used 2000 initial configurations for each nonequilibrium
MD simulation (if not stated otherwise for specific cases).
% They server as starting point of the branching
% trajectories.
The branching trajectories were integrated by the
Leap-frog scheme with the aforementioned nonequilibrium
temperature control technique.  The
time step was $\Delta t = 0.002~\textsf{ps}$. The short-range
interaction was smoothed from $0.8$ to $1.0~\textsf{nm}$ by the
``switch'' method provided by Gromacs.  An energy conserving PME
method was applied to calculate the electrostatic interaction in this
periodic system. In the thermostated region, the original dynamics was
coupled to a Langevin thermostat with $\tau_T = 0.1~\textsf{ps}$.
The whole system is also coupled to a Parrinello-Rahman barostat with $\tau_P = 2.0~\textsf{ps}$ to keep
the pressure at the ambient condition (1~Bar). Since the
change of the system size is small and slow, the pressure control
does not have an sizeable effect on the
dynamics of the system.

@@

\begin{figure}
  \centering
  \raisebox{0.2\height}{\includegraphics[width=0.22\textwidth]{c-2.eps}}
  \includegraphics[width=0.37\textwidth]{fig-begin-1.eps}
  \includegraphics[width=0.37\textwidth]{fig-end-1.eps}
  \caption{The probability of the conformations of the alanine dipeptide plotted on the Ramachandran histogram.
    (a): the
    equilibrium state at $t=0$~\textsf{ps}. (b): The new equilibrium
    state at $t=1000$~\textsf{ps}, i.e., fully relaxed under the
    action of a constant EF of 1~V/nm. The darker color
    indicates higher probability.}
  \label{fig:tmp4}
\end{figure}

@@

\begin{thebibliography}{10}

\bibitem{bohr2000microwave}
Henrik Bohr and Jakob Bohr.
\newblock Microwave-enhanced folding and denaturation of globular proteins.
\newblock {\em Physical Review E}, 61(4):4310, 2000.

\bibitem{bohr2000microwave-1}
Henrik Bohr and Jakob Bohr.
\newblock Microwave enhanced kinetics observed in ord studies of a protein.
\newblock {\em Bioelectromagnetics}, 21(1):68--72, 2000.

\bibitem{dePomerai2000cell}
David de~Pomerai, Clare Daniells, Helen David, Joanna Allan, Ian Duce, Mohammed
  Mutwakil, David Thomas, Phillip Sewell, John Tattersall, Don Jones, and Peter
  Candido.
\newblock Cell biology: Non-thermal heat-shock response to microwaves.
\newblock {\em Nature}, 405:417--418, 2000.

\bibitem{inskip2001cellular}
Peter~D Inskip, Robert~E Tarone, Elizabeth~E Hatch, Timothy~C Wilcosky,
  William~R Shapiro, Robert~G Selker, Howard~A Fine, Peter~M Black, Jay~S
  Loeffler, and Martha~S Linet.
\newblock Cellular-telephone use and brain tumors.
\newblock {\em New England Journal of Medicine}, 344(2):79--86, 2001.

\bibitem{mancinelli2004non}
Fabrizio Mancinelli, Michele Caraglia, Alberto Abbruzzese, Guglielmo
  d'Ambrosio, Rita Massa, and Ettore Bismuto.
\newblock Non-thermal effects of electromagnetic fields at mobile phone
  frequency on the refolding of an intracellular protein: Myoglobin.
\newblock {\em Journal of cellular biochemistry}, 93(1):188--196, 2004.

\bibitem{budi2005electric}
A.~Budi, F.S. Legge, H.~Treutlein, and I.~Yarovsky.
\newblock Electric field effects on insulin chain-b conformation.
\newblock {\em The Journal of Physical Chemistry B}, 109(47):22641--22648,
  2005.

\bibitem{budi2007effect}
A.~Budi, F.S. Legge, H.~Treutlein, and I.~Yarovsky.
\newblock Effect of frequency on insulin response to electric field stress.
\newblock {\em The Journal of Physical Chemistry B}, 111(20):5748--5756, 2007.

\bibitem{budi2008comparative}
Akin Budi, F~Sue Legge, Herbert Treutlein, and Irene Yarovsky.
\newblock Comparative study of insulin chain-b in isolated and monomeric
  environments under external stress.
\newblock {\em The Journal of Physical Chemistry B}, 112(26):7916--7924, 2008.

\bibitem{toschi2008effects}
Francesca Toschi, Francesca Lugli, Fabio Biscarini, and Francesco Zerbetto.
\newblock Effects of electric field stress on a $\beta$-amyloid peptide.
\newblock {\em The Journal of Physical Chemistry B}, 113(1):369--376, 2008.

\bibitem{astrakas2011electric}
Loukas Astrakas, Christos Gousias, and Margaret Tzaphlidou.
\newblock Electric field effects on chignolin conformation.
\newblock {\em Journal of Applied Physics}, 109(9):094702--094702, 2011.

\bibitem{astrakas2012structural}
Loukas~G Astrakas, Christos Gousias, and Margaret Tzaphlidou.
\newblock Structural destabilization of chignolin under the influence of
  oscillating electric fields.
\newblock {\em Journal of Applied Physics}, 111(7):074702--074702, 2012.

\bibitem{damm2012can}
Markus Damm, Christoph Nusshold, David Cantillo, Gerald~N Rechberger, Karl
  Gruber, Wolfgang Sattler, and C~Oliver Kappe.
\newblock Can electromagnetic fields influence the structure and enzymatic
  digest of proteins? a critical evaluation of microwave-assisted proteomics
  protocols.
\newblock {\em Journal of proteomics}, 75:5533--5543, 2012.

\bibitem{starzyk2013proteins}
A.~Starzyk and M.~Cieplak.
\newblock Proteins in the electric field near the surface of mica.
\newblock {\em The Journal of Chemical Physics}, 139:045102, 2013.

\bibitem{english2009nonequilibrium}
N.J. English, G.Y. Solomentsev, and P.~O'Brien.
\newblock Nonequilibrium molecular dynamics study of electric and low-frequency
  microwave fields on hen egg white lysozyme.
\newblock {\em The Journal of chemical physics}, 131:035106, 2009.

\bibitem{solomentsev2012effects}
G.Y. Solomentsev, N.J. English, and D.A. Mooney.
\newblock Effects of external electromagnetic fields on the conformational
  sampling of a short alanine peptide.
\newblock {\em Journal of Computational Chemistry}, 33:917--923, 2012.

>>>>

The possible effect of the electric field (EF) on the conformation of
proteins is a subject of increasing attention since it is linked to the problem of exposing human tissues to
electromagnetic radiation. Despite its major relevance, a
satisfactory understanding of how the EF influences the behaviour of proteins
has not been reached so far although a considerable number of
experimental e.g.~\cite{bohr2000microwave, bohr2000microwave-1,
  dePomerai2000cell, inskip2001cellular, mancinelli2004non} and
theoretical e.g.~\cite{budi2005electric, budi2007effect,
  budi2008comparative, toschi2008effects, astrakas2011electric,
  astrakas2012structural, damm2012can, starzyk2013proteins,
  english2009nonequilibrium, solomentsev2012effects}
studies have been devoted to this topic.

In this perspective, molecular dynamics (MD) has been proven useful in
understanding the behavior of a protein 
in both static and oscillatory EF~\cite{budi2005electric, budi2007effect, budi2008comparative,
  toschi2008effects, astrakas2011electric, astrakas2012structural,
  damm2012can, starzyk2013proteins, english2009nonequilibrium,
  solomentsev2012effects}. In these studies, the initial
configurations are usually taken from the Protein Data Bank (PDB),
and then temperature controlled simulations are performed for a maximum
time range of the order of 10-60 nanoseconds.
During this simulation time 
relevant conformational changes characterizing the solvated molecule can be observed. 
Along the trajectories, various observable, such as the
secondary structure, root mean square displacement (RMSD), the dipole
moment, shape parameters are calculated as a function of time, or
in stationary condition, taking time averages.
This approach releases useful information on how the
secondary structure of the molecule is changed by external  EF.  However,
the behavior of observables in time is characterized by 
strong
fluctuations. The entity of the fluctuations is such that a quantitative 
analysis becomes difficult and does not allow for  a clear understanding of the rationale behind the conformational changes.
A time average can reduce the statistical uncertainty,
but in non-stationary conditions, it is a not-well-founded procedure.
In fact, we are studying non-stationary nonequilibrium process
(e.g. relaxing process to the new conformations under a static
EF, or the dynamics from one conformation to the other under an oscillatory EF).
In this perspective, if one wants to use MD, the proper way to do it would be that of treating a situation of nonequilibrium with algorithms that can explicitly describe it.
This is the approach that will be followed in this work, using
the dynamical nonequilibrium molecular dynamics simulation (D-NEMD).
The method, firstly developed by one of us and coworkers~\cite{ciccotti1975direct, ciccotti1979thought}, 
has been recently applied, to the study of
hydrodynamics~\cite{orlandini2011hydrodynamics,
  orlandini2011hydrodynamics-01}. The essential feature of D-NEMD is to
provide a way of evaluating time-dependent
nonequilibrium averages starting from a properly determined initial ensemble.
In this context, the method can be used to analyze the
conformational changes of a solvated molecule, subject to a time-dependent EF.
In the present paper, we study a small peptide, alanine dipeptide, as a model for showing how the method can be used. 
The choice of a small peptide, rather than a larger one, is done firstly because it
is easier to draw clear conclusions on the nonequilibrium properties
of a single peptide by excluding the slow-going interplay between
different peptide segments along a chain.
Secondly, the computational cost required is rather massive and thus  testing
the various methodology proposed here would be hampered for large molecules,
although they remain our goal.

The paper is organized as follows:
We firstly describe the idea of D-NEMD, then
introduce the physical
observables chosen to analyze the nonequilibrium responses.
The numerical results for the response of alanine
dipeptide under static and oscillatory EF are
finally presented. By calculating the nonequilibrium properties from D-NEMD
simulations, we are able to address
a few relavant questions: (i)~How the conformation of the peptide
changes in an EF (conformational space);
(ii)~When do these changes happen (time scale),
and how likely they are (probability). 
To our knowledge, this study is one of the few (if not the first)  where
time-dependent nonequilibrium w.r.t. conformation
is explicitly considered for a large molecule
in solution subject to time-dependent EF.

\section{Methodology}
\subsection{Nonequilibrium molecular dynamics simulation}
In this section we briefly review the dynamical approach to nonequilibrium
molecular dynamics (D-NEMD) \cite{ciccotti1975direct, ciccotti1979thought,
  orlandini2011hydrodynamics, orlandini2011hydrodynamics-01}.
In the following we denote the macroscopic observable by $O(t)$. If at $t$
the configurational probability distribution is $\rho(\vect x, t)$, where
$\vect x$ is the phase space variable, then the observable can be
expressed by
\begin{align}\label{eqn:tmp1}
  O(t) = \int d\vect x\, \hat O(\vect x)\rho(\vect x, t)  = \langle \hat O(\vect x), \rho(\vect x, t)\rangle,
\end{align}
we refer to $\hat O (\vect x)$ as the microscopic observable, 
which is measured at each point in the phase space $\vect x$.
Here we always assume that the initial probability distribution
$\rho(\vect x, 0)$ is known. In particular, in our case,
it is identical to the equilibrium distribution of the system
without EF.
The bracket on the right hand side of Eq.~\eqref{eqn:tmp1} denotes, as usual, the inner product in the
phase space.  We assume that he dynamics of the system is governed by the
Hamiltonian equation, i.e. $\dot {\vect x} = J \cdot \nabla_{\vect x}
\mathcal H(\vect x, t)$, where $\mathcal H$ is the time-dependent Hamiltonian,
and $J$ is
the symplectic matrix. The Liouville equation for the probability
distribution writes:
\begin{align}\label{eqn:tmp2}
  \frac{\partial \rho(\vect x, t)}{\partial t} = - iL(t) \rho(\vect x, t),
\end{align}
where $iL(t) = \{\cdot, \mathcal H\}$ is the Liouville operator.
Eq.~\eqref{eqn:tmp2}
can be formally solved by $\rho(\vect x, t) = U^\dagger(t,0) \rho(\vect x, 0)$,
where $U^\dagger(t,0) = \exp\{-i \int_0^t d t' L(t')\}$.
On the other hand
\begin{align}
  \frac{d \hat O(\vect x(t))}{dt} = \nabla_{\vect x}\hat O\cdot \dot{\vect x}
  = \nabla_{\vect x}\hat O\cdot J\cdot \nabla_{\vect x}\mathcal H
  = iL(t) \hat O (\vect x(t))
\end{align}
This equation can be formally solved by
$\hat O(\vect x(t)) = U(t,0)\,\hat O(\vect x(0))$, therefore,
\begin{align}\nonumber
  O(t) & = \langle \hat O(\vect x), \rho(\vect x, t)\rangle
  = \langle \hat O(\vect x), U^\dagger(t,0)\, \rho(\vect x, 0)\rangle
  = \langle U(t,0)\,\hat O(\vect x), \rho(\vect x, 0)\rangle\\\label{eqn:tmp4}
  &= \langle \hat O(\vect x(t)), \rho(\vect x, 0)\rangle
\end{align}
Since we assume that the system starts from the distribution of equilibrium (without EF)
 Eq.~\eqref{eqn:tmp4} expresses the fact that the 
observable $O(t)$ calculated in a situation of nonequilibrium is equal to the ensemble average of
the microscopic observable computed along trajectories, starting from
initial configurations sampled from an \emph{equilibrium}
distribution. In practical terms we proceed by first running an equilibrium MD simulation
to generate a sample of configurations. Next we employ these configurations as
initial configurations and, for each, the full Hamiltonian dynamics is integrated until
time $t$.
In this paper, these trajectories are called \emph{branching
  trajectories} (see Fig.\ref{branching} for a pictorial representation). 
Finally, the value of the macroscopic observable at time $t$ is
calculated by averaging the microscopic observable measured at each time $t$ along each branching trajectory.
Stochastic dynamics (e.g.~Langevin dynamics) can be handled analogoursly.

@@

\subsection{Control of temperature in nonequilibrium MD}\label{sec:tmp2b}

\begin{figure}
  \centering
  \includegraphics[width=0.3\textwidth]{thermostat.eps}
  \caption{Schematic plot of the nonequilibrium temperature control scheme.
    The white area is the region characterized by pure dynamics without any direct action by the external thermostat.
    The gray area is the region where the thermostat is used to keep the target temperature.
    $R_{ex}$ denotes the size of the purely dynamical region, which
    is a sphere around the molecule of interest (in our case it is the alanine
    dipeptide).
    % at the center of the box.
    The whole system is subject to periodic boundary conditions. }
  \label{fig:tmp2}
\end{figure}

% The aforementioned D-NEMD simulation algorithm
% does not describe how the boundary of the system should be set up.
% The very naive idea of doing the simulation
% is to simulate a infinitely large isolated system, or any large enough
% isolated subsystem of it.
% In practice, 
% Therefore, we are interested in a 
For this study, we are interested in 
nonthermal~\cite{delaHoz2005microwaves} effects of the EF
on the configuration of the solvated alanine dipeptide.
For this reason a proper control of the temperature in a nonequilibrium system
becomes an important issue; in fact it was argued that  an
accurate control of the internal reaction temperature is essential to
perform reproducible experiments~\cite{damm2012can}.
A straightforward (and realistic) implementation of the temperature control consists in 
embedding the alanine into an infinitely large solvent environment, and 
the EF is only applied to the neighborhood of the alanine. In this case
any extra heat generated by the EF can be effectively
dispersed in the large solvent environment.
In this way one can identify the effect of the EF on the conformational space of the solvated molecule,
without the artifact that the surrounding solvent cannot release the additional energy of the EF.
However, in practice, in most of the cases the computational cost of having large solvent environment is rather massive if not prohibitive.
Therefore, it is of interest to substitute the infinitely large
solvent environment with a smaller affordable system and at the same time accurately reproduce the nonequilibrium
physical properties of interest.
In MD, for equilibrium situations, the solution is to couple
the system to an external thermostat, under which the desired ensemble
(canonical in this case) can be
sampled by simulating a finite size system with periodic boundary
conditions.
However, coupling the system to a thermostat for nonequilibrium situations is a more delicate issue.
In fact, since a thermostat is explicitly designed to bring the system into
thermal equilibrium,
the perturbation produced by the EF would be mostly adsorbed by the thermostat. This means that the physics of the original dynamics in presence of the EF
would be lost and the response to the perturbation would be unphysical. 
%will disturb the original Hamiltonian dynamics. This is
%in general problematic, 
%because the dynamics  under thermostating is artificial designed,
%so that the equilibrium distribution can be correctly produce, but not
%the nonequilibrium properties of system.
% dynamics of the system is no longer Hamiltonian, and is
% artificial designed to reach the equilibrium probability distribution.
% However, for a nonequilibrium system, the \emph{dynamics} properties
% will be distrubed by the thermostat, because the
% dynamics of the system is no longer Hamiltonian, and is
% artificial designed to reach the equilibrium probability distribution.
% For a nonequilibrium simulation, simply couple the system
% to a thermostat may not be a good idea, because the \emph{dynamics}
% generated by the thermostat is artificial, and deviation from the
% Hamiltonian dynamics under investigation will introduce artificial
% a effects in the nonequilibrium observations.

In this section we propose a procedure to solve the problem: we
start observing that the dynamics without thermostat of the alanine, 
and the water molecules in the closer solvation shells, is
crucial to determine the response of the observable of interest
under the EF perturbation
(in this work, the observable is the conformational change of alanine).
Instead the detailed dynamics of
the water molecules far away is not of primary importance: the region beyond the first solvation shells plays mainly the role of a thermodynamic bath. 
Therefore, instead of simulating an infinitely
large system, we simulate our finite size system with periodic
boundary condition, and with the following characteristics:
(i)~The simulating region is divided in two
subregions, a spherical region around the alanine dipeptide of radius $R_{ex}$,
centered at the alpha-carbon. Here the dynamics is
not subject to a thermostat.
We refer to such subregion as \emph{the  dynamical
region}. (ii)~Beyond $R_{ex}$, the dynamics of the water molecules is coupled
to a Langevin thermostat.  This region is called \emph{the thermostated
region}. For technical convenience we
fix the position of the alpha-carbon in space, so that the alanine is always
located at the center of the simulation region (see Fig.~\ref{fig:tmp2}).
In these conditions, the  dynamics is preserved in the dynamical region (where the properties of interest are observed), while 
possible artificial effects of the thermostating process in the outside region
are negligible due to the finite correlation
range of liquid water.
At the same time, the thermostated region works as a infinitely large
environment that effectively absorbs the extra heat produced by the EF.
The validity of the above
statement will be checked by a series of numerical tests where it is shown that
the response of observables in nonequilibrium (i.e. under the effect of the EF) does not depend on the size of the system
and on the size of the  dynamical subregion, provided that they are
reasonably large.

@@

\section{Case I: Alanine dipeptide
  under a uniform  constant EF}

\subsection{System settings and simulation protocol}
The system is set up in a $2.7\times 2.7\times
2.7\, \textsf{nm}^3$ periodic simulation region, with one alanine dipeptide
described by the CHARMM27 force field, and dissolved in 644 TIP3P
water molecules.
%The EF exserts force $\vect F = q \vect E(t)$ on
%the partial charge $q$ in the system.
The size of the  dynamical region is $R_{ex} = 1.0$~nm.
All simulations are performed by a home-modified GROMACS 4.5~\cite{pronk2013gromacs}.
First, an equilibrium NVT simulation of
100~\textsf{ns} was performed with a Langevin thermostat (time-scale
$\tau_T = 0.5~\textsf{ps}$).  Along the trajectory, configurations were taken every
50~\textsf{ps} and we used 2000 initial configurations for each nonequilibrium
MD simulation (if not stated otherwise for specific cases).
% They server as starting point of the branching
% trajectories.
The branching trajectories were integrated by the
Leap-frog scheme with the aforementioned nonequilibrium
temperature control technique.  The
time step was $\Delta t = 0.002~\textsf{ps}$. The short-range
interaction was smoothed from $0.8$ to $1.0~\textsf{nm}$ by the
``switch'' method provided by Gromacs.  An energy conserving PME
method was applied to calculate the electrostatic interaction in this
periodic system. In the thermostated region, the original dynamics was
coupled to a Langevin thermostat with $\tau_T = 0.1~\textsf{ps}$.
The whole system is also coupled to a Parrinello-Rahman barostat~\cite{parrinello1981polymorphic} with $\tau_P = 2.0~\textsf{ps}$ to keep
the pressure at ambient condition (1~Bar). Since the
change of the system size is small and slow, the pressure control
does not have an sizeable effect on the
dynamics of the system.
At time $t=0$~ps, the system
has been fully equilibrated without any EF. From $t=0$ to
the warm-up time $t=t_{\textrm{warm}}$, the EF is switched on linearly, while
after $t=t_{\textrm{warm}}$, the field is kept constant in time at
$E_{\infty}$.  In this work we consider $t_{\textrm{warm}} = 10$~ps,
and $E_{\infty} = 1$~V/nm.
The direction of the field is arbitrarily chosen along the
$x$ direction.
In this paper we denote the
EF as a function of time $\vect E(t)$.
Therefore, for the case of constant EF, we have
$\vect E(t) = (E_\infty\cdot t/t_{warm},0,0)$ for $0\leq t < t_{warm}$, and 
$\vect E(t) = (E_\infty,0,0)$
for $t \geq t_{warm}$.

@@

\begin{figure}
  \centering
  \raisebox{0.2\height}{\includegraphics[width=0.22\textwidth]{c-2.eps}}
  \includegraphics[width=0.37\textwidth]{fig-begin-1.eps}
  \includegraphics[width=0.37\textwidth]{fig-end-1.eps}
  \caption{The probability of the conformations of the alanine dipeptide plotted on the Ramachandran histogram.
    (a): the
    equilibrium state at $t=0$~\textsf{ps}. (b): The new equilibrium
    state at $t=1000$~\textsf{ps}, i.e., fully relaxed under the
    action of a constant EF of 1~V/nm.  Darker color
    indicates higher probability.}
  \label{fig:tmp4}
\end{figure}

@@

\bibliography{ref}{}
\bibliographystyle{unsrt}
